---
title: "Mastodon Data"
date: "`r Sys.Date()`"
output:
  rmdformats::material:
    highlight: kate
    self_contained: no
    lib_dir: "assets"
---

# Data Overview

```{r knitr_init, echo=FALSE, cache=FALSE, include=FALSE}
library(knitr)
library(rmdformats)
library(DT)
library(tidyverse)
library(stringr)
library(purrr)
library(lubridate)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r setup}
mastodon <- read_rds("data/mastodon.rds")
```

## Tracked instances

Instances are scraped from [the list at  the tootsuite/mastodon repo](https://github.com/tootsuite/mastodon/blob/master/docs/Using-Mastodon/List-of-Mastodon-instances.md), if your instance is missing, please add it to the listed document.

```{r instance_list}
mastodon %>%
  group_by(instance) %>%
  summarise(n = n(), users_last = last(users), 
            data_first = min(date), data_last = max(date)) %>%
  arrange(desc(users_last)) %>%
  rename(Instance = instance,
         Data_Points = n,
         Users_Last = users_last,
         First_Data = data_first,
         Last_Data = data_last) %>%
  datatable()
```

# About

This is just a quick demo proof of concept by @jemus42@mastodon.social / @jemus42 on twitter.  
A more detailed analysis will come, and I'll probably scrape data daily for long-term analysis.

The site is built in Rmarkdown, using the [material theme from juba/rmdformats](https://github.com/juba/rmdformats#material-format).

The code for this project is [on GitHub](https://github.com/jemus42/mastodon-instance-stats).
